---
title: "State Level Analysis of Risk for Asthma and Covid-19"
author: "Esteban Correa"
date: "April 21, 2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
library(tidyverse)
library(SpatialEpi)
library(raster)
library(rgeos)
library(rgdal)
library(sf)
library(leaflet)
library(RColorBrewer)
library(ggthemes)
library(ggplot2)
library(lubridate)
library(classInt)
library(brinla)
library(INLA)
library(spdep)
covid19<-read.csv("../covid19/data/jhucounty04-20-2020.csv") %>%
  dplyr::select(FIPS,Covid.Confirmed,Covid.Deaths)
asthmaDeaths<-read.csv("../asthma/CDCWONDER/asthmaDeaths.csv") %>%
  dplyr::select(County.Code,asthma.Deaths,asthma.Population,Crude.Rate) 
cdi<-read.csv("../covid19/data/SVI2018/CDIByState.csv")
pm25<-read.csv("pm2.5byCounty.csv")
demog<-read.csv("../asthma/factFinder/ACS5Years2018.csv")
# load("pm25STCounty.RData")
uscounty<-readOGR("../covid19/data/SVI2018","SVI2018_US_county")
uscounty@data<-uscounty@data %>%
  select(FIPS,ST,ST_ABBR,E_TOTPOP,EP_POV)

# ICD-10 Codes: J40-J47 (Chronic lower respiratory diseases)
```

# Covid19 at April 20-2020, Asthma deaths from 2010-2017

```{r,include=FALSE}
asthmaDeaths$asthma.Deaths<-as.numeric(as.character(asthmaDeaths$asthma.Deaths))
asthmaDeaths$asthma.Population<-as.numeric(as.character(asthmaDeaths$asthma.Population))
asthmaDeaths$Crude.Rate<-as.numeric(as.character(asthmaDeaths$Crude.Rate))
asthmaDeaths$fips<-sprintf("%05d", asthmaDeaths$County.Code)
asthmaDeaths$asthma.Deaths[is.na(asthmaDeaths$asthma.Deaths)]<-0

covid19$id<-sprintf("%05d", covid19$FIPS) 
covid19$Covid.Deaths[is.na(covid19$Covid.Deaths)]<-0


demog$fips<-substring(demog$GEO_ID,10,14)
demog<-demog%>% 
  select(fips,DP05_0037PE,DP05_0038PE,DP05_0071PE,DP05_0002PE,DP05_0003PE,DP05_0024PE)
colnames(demog)<-c("fips","whitePE","blackPE","latinoPE","malePE","femalePE","over65PE")

pm25$fips<-sprintf("%05d", pm25$fips)
pm25$pm25avg<-rowMeans(pm25[,2:20],na.rm = T)
pm25<-pm25%>% 
  select(fips,pm25avg)



```



<!-- # County level PM 2.5 -->

```{r,include=FALSE}
uscounty <- merge(uscounty, cdi,by.x="ST_ABBR",by.y="ST_ABBR") #Chronic diseases
uscounty <- merge(uscounty, covid19,by.x="FIPS",by.y="id") #covidDea
uscounty <- merge(uscounty, asthmaDeaths,by.x="FIPS",by.y="fips") #asthmaDe
uscounty <- merge(uscounty, demog,by.x="FIPS",by.y="fips") #demog
uscounty <- merge(uscounty, pm25,by.x="FIPS",by.y="fips") #pm25

uscounty$asthma.Deaths[is.na(uscounty$asthma.Deaths)]<-0
uscounty$Covid.Deaths[is.na(uscounty$Covid.Deaths)]<-0
#removing alaska y puerto rico y NY
uscounty<-uscounty[uscounty$ST!="02",]
uscounty<-uscounty[uscounty$ST!="15",]
uscounty<-uscounty[uscounty$ST!="36",]

uscounty<-uscounty[uscounty$FIPS!="35039",]
head(uscounty,100)
```




```{r,include=FALSE}

# head(uscounty,100)
uscounty$E_TOTPOP<-as.numeric(as.character(uscounty$E_TOTPOP))
uscounty$Easthma <- expected(uscounty$E_TOTPOP, uscounty$asthma.Deaths, 1)
uscounty$Ecovid <- expected(uscounty$E_TOTPOP, uscounty$Covid.Deaths, 1)

uscounty$SIR.asthma <- uscounty$asthma.Deaths/uscounty$Easthma
uscounty$SIR.covid <- uscounty$Covid.Deaths/uscounty$Ecovid
summary(uscounty)
```

# Covariates

EP_MINRTY = Percentage minority (all persons except white, nonHispanic) estimate E_PCI Per capita income estimate, 2014-2018 ACS.
EP_CROWD = At household level (occupied housing units), more people than rooms estimate, 2014-2018 ACS
EP_AGE65  = percentage of people age 65 over
EP_POV = Percentage of people living under poverty
EP_CROWD = Percentage of occupied housing units with more people than rooms estimate

<!-- # Asthma deaths 1 -->

<!-- $Asthma_{Deaths}\sim covidMortality_{100k} + pm25_{2018} + EP\_POV_{\%} + EP\_AGE65_{\%} + EP\_MINRTY_{\%} + EP\_CROWD_{\%} + (1|STATE)$ -->

<!-- ```{r,include = FALSE} -->
<!-- sirMap<-uscounty -->
<!-- sirMap$E_PCI<-as.numeric(sirMap$E_PCI) -->
<!-- sirMap$E_MINRTY<-as.numeric(sirMap$E_MINRTY) -->
<!-- sirMap$EP_AGE65<-as.numeric(sirMap$EP_AGE65) -->
<!-- sirMap$Covid.Confirmed<-as.numeric(sirMap$Covid.Confirmed) -->
<!-- # hist(sirMap$Covid.Confirmed,128) -->
<!-- sirMap$Covid.Confirmed.p100k<-sirMap$Covid.Confirmed*100/sirMap$E_TOTPOP -->
<!-- sirMap$Covid.Deaths.p100k<-sirMap$Covid.Deaths*100/sirMap$E_TOTPOP -->
<!-- sirMap$asthma.Deaths.p100k<-sirMap$asthma.Deaths*100/sirMap$E_TOTPOP -->

<!-- sirMap<-sirMap[sirMap$EP_POV>=0,] -->
<!-- sirMap<-sirMap[sirMap$EP_AGE65>=0,] -->
<!-- sirMap<-sirMap[sirMap$EP_MINRTY>=0,] -->
<!-- sirMap<-sirMap[sirMap$EP_CROWD>=0,] -->
<!-- # nb <- poly2nb(sirMap) -->
<!-- # nb2INLA("us.adj", nb) -->
<!-- # g <- inla.read.graph(filename = "us.adj") -->
<!-- # sirMap$re_u <- 1:nrow(sirMap@data) -->
<!-- sirMap$re_v <- 1:nrow(sirMap@data) -->


<!-- f1<-asthma.Deaths ~  Covid.Deaths.p100k + pm25 + EP_POV + EP_AGE65 + EP_MINRTY + EP_CROWD + f(STATE,model = "iid") + f(re_v, model = "iid") -->
<!-- inla.m1 <- inla( f1, -->
<!--                  family = "poisson",data = sirMap@data,E = Easthma, -->
<!--                  num.threads = 8,verbose = F, -->
<!--                  control.predictor = list(compute = TRUE), -->
<!--                  control.compute = list(dic = TRUE,waic=T,cpo=TRUE)) -->

<!-- sirMap$RRasth <- inla.m1$summary.fitted.values[, "mean"] -->
<!-- sirMap$LLasth <- inla.m1$summary.fitted.values[, "0.025quant"] -->
<!-- sirMap$ULasth <- inla.m1$summary.fitted.values[, "0.975quant"] -->
<!-- ``` -->


<!-- ```{r,echo=FALSE} -->

<!-- summary(inla.m1) -->
<!-- round(exp(inla.m1$summary.fixed[,c(1,3,5)]),2) -->

<!-- # bri.Pois.resid(inla.m1) -->
<!-- #check random effects -->
<!-- # bri.hyperpar.summary(inla.m1) -->
<!-- # check poisson assumption where u(x)=var(x) -->
<!-- # bri.hyperpar.plot(inla.m1) -->


<!-- tab <- inla.m1$summary.random$STATE -->
<!-- names(tab)[c(4, 6)] <- c("IClower", "ICupper") -->
<!-- tab<-tab[order(tab$ID),] -->
<!-- tab$idx<-nrow(tab):1 -->
<!-- tab$mean_exp<-exp(tab$mean) -->
<!-- tab<-tab[!is.infinite(tab$mean_exp),] -->
<!-- tab$IClower_exp<-exp(tab$IClower) -->
<!-- tab$ICupper_exp<-exp(tab$ICupper) -->
<!-- ID<-tab$ID -->
<!-- ggplot(data = tab, aes(x = mean_exp, y = idx)) + -->
<!--   geom_point() + xlab(expression(u[p])) + -->
<!--   scale_x_continuous(breaks = NULL)+ -->
<!--   scale_y_continuous("STATE", breaks=nrow(tab):1,labels=c(ID)) + -->
<!--   geom_errorbarh(aes(xmin = IClower_exp, xmax = ICupper_exp) ) + -->
<!--   geom_vline (xintercept = 1, linetype = 2)+ -->
<!--   theme_fivethirtyeight() -->
<!-- ``` -->


<!-- ```{r,echo=FALSE} -->

<!-- brks<-c(min(sirMap$RRasth),1.0,max(sirMap$RRasth)) -->
<!-- ggplot() + -->
<!--   geom_sf(data=st_as_sf(sirMap),aes(fill = RRasth)) + -->
<!--   scale_fill_gradient2( -->
<!--     midpoint = 1, low = "blue", mid = "white", high = "red", -->
<!--      breaks = brks, -->
<!--     labels = round(brks,1) -->
<!--   ) + -->
<!--   ggtitle("Relative Risk Asthma risk in Covid-19 Pandemic")+ -->
<!--   coord_sf(crs = mycrs)+ -->
<!--   theme_minimal() -->

<!-- ``` -->

<!-- # Asthma deaths 2 -->

avgObesityPrev

$Asthma_{Deaths}\sim covidMortality_{100k} + pm25_{avg} + EP\_POV_{\%} + (1|STATE)$

```{r}
sirMap<-uscounty
# sirMap$E_PCI<-as.numeric(sirMap$E_PCI)
# sirMap$E_MINRTY<-as.numeric(sirMap$E_MINRTY)
# sirMap$EP_AGE65<-as.numeric(sirMap$EP_AGE65)
sirMap$Covid.Confirmed<-as.numeric(sirMap$Covid.Confirmed)
# hist(sirMap$Covid.Confirmed,128)
sirMap$Covid.Confirmed.p100k<-sirMap$Covid.Confirmed*100/sirMap$E_TOTPOP
sirMap$Covid.Deaths[is.na(sirMap$Covid.Deaths)]<-0
sirMap$Covid.Deaths.p100k<-sirMap$Covid.Deaths*100/sirMap$E_TOTPOP
sirMap$asthma.Deaths[is.na(sirMap$asthma.Deaths)]<-0
sirMap$asthma.Deaths.p100k<-sirMap$asthma.Deaths*100/sirMap$E_TOTPOP


# nb <- poly2nb(sirMap)
# nb2INLA("us.adj", nb)
# g <- inla.read.graph(filename = "us.adj")
# sirMap$re_u <- 1:nrow(sirMap@data)
sirMap$re_v <- 1:nrow(sirMap@data)

f1<-asthma.Deaths ~  Covid.Deaths + pm25avg + EP_POV + blackPE+latinoPE + femalePE + over65PE + avgObesityPrev + smokingPrev + f(ST_ABBR,avgObesityPrev,model = "iid") + f(ST_ABBR,smokingPrev,model = "iid") + f(re_v, model = "iid")
inla.m2 <- inla( f1,
                 family = "zeroinflatedpoisson1",
                 data = sirMap@data,E = Easthma,
                 num.threads = 8,verbose = F,
                 control.predictor = list(compute = TRUE),
                 control.compute = list(dic = TRUE,waic=T,cpo=TRUE))

sirMap$RRasth2 <- inla.m2$summary.fitted.values[, "mean"]
sirMap$LLasth2 <- inla.m2$summary.fitted.values[, "0.025quant"]
sirMap$ULasth2 <- inla.m2$summary.fitted.values[, "0.975quant"]

round(exp(inla.m2$summary.fixed[,c(1,3,5)]),2)
bri.hyperpar.plot(inla.m2)
bri.Pois.resid(inla.m2)
tab <- inla.m2$summary.random$ST_ABBR
names(tab)[c(4, 6)] <- c("IClower", "ICupper")
tab<-tab[order(tab$ID),]
tab$idx<-nrow(tab):1
tab$mean_exp<-exp(tab$mean)
tab<-tab[!is.infinite(tab$mean_exp),]
tab$IClower_exp<-exp(tab$IClower)
tab$ICupper_exp<-exp(tab$ICupper)
ID<-tab$ID
ggplot(data = tab, aes(x = mean_exp, y = idx)) +
  geom_point() + xlab(expression(u[p])) +
  scale_x_continuous(breaks = NULL)+
  scale_y_continuous("STATE", breaks=nrow(tab):1,labels=c(ID)) +
  geom_errorbarh(aes(xmin = IClower_exp, xmax = ICupper_exp) ) +
  geom_vline (xintercept = 1, linetype = 2)+
  theme_fivethirtyeight()
```

# brms

```{r}
library(brms)
library(brinla)
```

```{r}
brmsDF<-as.data.frame(sirMap)

brm.fl1 <- brm(asthma.Deaths ~ offset(log(Easthma)) + Covid.Deaths + pm25avg + EP_POV + blackPE + latinoPE + femalePE + over65PE + avgObesityPrev + smokingPrev + (1+avgObesityPrev + smokingPrev|ST_ABBR), 
              data = brmsDF, 
              family = "zero_inflated_poisson",
              # family = poisson(),
              iter = 3000, cores  = 8,seed = 2020)  
summary(brm.fl1)
round(exp(summary(brm.fl1)$fixed[,c(1,3,4)]),2)

pp_check(brm.fl1,nsamples = nrow(sirMap)/1)
plot(brm.fl1, N = 2, ask = FALSE)
plot(brm.fl1, points = TRUE,ask=FALSE)
# names(brm.bmi)
# ## extract fitted values
fitted_values <- fitted(brm.fl1)
```



```{r}
brmsDF<-as.data.frame(sirMap)

brm.fl2 <- brm(Covid.Deaths ~ offset(log(Ecovid)) + asthma.Deaths + pm25avg + EP_POV + blackPE + latinoPE + femalePE + over65PE +avgObesityPrev + smokingPrev + (1+avgObesityPrev + smokingPrev|ST_ABBR), 
              data = brmsDF, 
              family = "zero_inflated_poisson",
              # family = poisson(),
              iter = 3000, cores  = 8,seed = 2020)  
summary(brm.fl2)
round(exp(summary(brm.fl2)$fixed[,c(1,3,4)]),2)

pp_check(brm.fl2,nsamples = nrow(brmsDF)/1)
plot(brm.fl2, N = 2, ask = FALSE)
plot(brm.fl2, points = TRUE,ask=FALSE)
# names(brm.bmi)
# ## extract fitted values
fitted_values <- fitted(brm.fl1)
```




# Covid deaths

$Covid_{Deaths}\sim asthmaMortality_{100k} + pm25_{2018} + EP\_POV_{\%} + EP\_AGE65_{\%} + EP\_MINRTY_{\%} + EP\_CROWD_{\%} + (1|STATE)$

<!-- ```{r,include=FALSE} -->

<!-- # plot(sirMap$asthma.Deaths.rate) -->

<!-- f1<- Covid.Deaths ~  sirMap$asthma.Deaths.p100k + pm25avg + EP_POV + EP_AGE65 + EP_MINRTY + EP_CROWD + f(STATE,model = "iid") + f(re_v, model = "iid") -->
<!-- inla.m1 <- inla( f1, -->
<!--                  family = "zeroinflatedpoisson1", -->
<!--                  data = sirMap@data,E = Easthma, -->
<!--                  num.threads = 8,verbose = F, -->
<!--                  control.predictor = list(compute = TRUE), -->
<!--                  control.compute = list(dic = TRUE,waic=T,cpo=TRUE)) -->

<!-- sirMap$RRcov <- inla.m1$summary.fitted.values[, "mean"] -->
<!-- sirMap$LLcov <- inla.m1$summary.fitted.values[, "0.025quant"] -->
<!-- sirMap$ULcov <- inla.m1$summary.fitted.values[, "0.975quant"] -->
<!-- ``` -->


```{r}
sirMap<-uscounty

sirMap$Covid.Confirmed<-as.numeric(sirMap$Covid.Confirmed)
sirMap$Covid.Confirmed.p100k<-sirMap$Covid.Confirmed*100/sirMap$E_TOTPOP
sirMap$Covid.Deaths[is.na(sirMap$Covid.Deaths)]<-0
sirMap$Covid.Deaths.p100k<-sirMap$Covid.Deaths*100/sirMap$E_TOTPOP
sirMap$asthma.Deaths[is.na(sirMap$asthma.Deaths)]<-0
sirMap$asthma.Deaths.p100k<-sirMap$asthma.Deaths*100/sirMap$E_TOTPOP

sirMap$re_v <- 1:nrow(sirMap@data)

f1<-Covid.Deaths ~  asthma.Deaths + pm25avg + EP_POV + blackPE+latinoPE + femalePE + over65PE + avgObesityPrev + smokingPrev + diabetesPrev + ckdPrev + f(ST_ABBR,model = "iid") + f(re_v, model = "iid")
inla.m2 <- inla( f1,
                 family = "zeroinflatedpoisson1",
                 data = sirMap@data,E = Ecovid,
                 num.threads = 8,verbose = F,
                 control.predictor = list(compute = TRUE),
                 control.compute = list(dic = TRUE,waic=T,cpo=TRUE))

sirMap$RRasth2 <- inla.m2$summary.fitted.values[, "mean"]
sirMap$LLasth2 <- inla.m2$summary.fitted.values[, "0.025quant"]
sirMap$ULasth2 <- inla.m2$summary.fitted.values[, "0.975quant"]

round(exp(inla.m2$summary.fixed[,c(1,3,5)]),2)
bri.hyperpar.plot(inla.m2)
bri.Pois.resid(inla.m2)

tab <- inla.m2$summary.random$ST_ABBR
names(tab)[c(4, 6)] <- c("IClower", "ICupper")
tab<-tab[order(tab$ID),]
tab$idx<-nrow(tab):1
tab$mean_exp<-exp(tab$mean)
tab<-tab[!is.infinite(tab$mean_exp),]
tab$IClower_exp<-exp(tab$IClower)
tab$ICupper_exp<-exp(tab$ICupper)
ID<-tab$ID
ggplot(data = tab, aes(x = mean_exp, y = idx)) +
  geom_point() + xlab(expression(u[p])) +
  scale_x_continuous(breaks = NULL)+
  scale_y_continuous("STATE", breaks=nrow(tab):1,labels=c(ID)) +
  geom_errorbarh(aes(xmin = IClower_exp, xmax = ICupper_exp) ) +
  geom_vline (xintercept = 1, linetype = 2)+
  theme_fivethirtyeight()
```



```{r,echo=FALSE}
round(exp(inla.m1$summary.fixed[,c(1,3,5)]),2)

tab <- inla.m1$summary.random$STATE
names(tab)[c(4, 6)] <- c("IClower", "ICupper")
tab<-tab[order(tab$ID),]
tab$idx<-nrow(tab):1
tab$mean_exp<-exp(tab$mean)
tab<-tab[!is.infinite(tab$mean_exp),]
tab$IClower_exp<-exp(tab$IClower)
tab$ICupper_exp<-exp(tab$ICupper)
ID<-tab$ID
ggplot(data = tab, aes(x = mean_exp, y = idx)) +
  geom_point() + xlab(expression(u[p])) +
  scale_x_continuous(breaks = NULL)+
  scale_y_continuous("STATE", breaks=nrow(tab):1,labels=c(ID)) +
  geom_errorbarh(aes(xmin = IClower_exp, xmax = ICupper_exp) ) +
  geom_vline (xintercept = 1, linetype = 2)+
  theme_fivethirtyeight()

```


```{r,echo=FALSE}

brks<-c(min(sirMap$RRcov),1.0,max(sirMap$RRcov))
ggplot() +
  geom_sf(data=st_as_sf(sirMap),aes(fill = RRcov)) +
  scale_fill_gradient2(
    midpoint = 1, low = "blue", mid = "white", high = "red",
     breaks = brks,
    labels = round(brks,1)
  ) +
  ggtitle("Relative Risk for Covid-19 Pandemic")+
  coord_sf(crs = mycrs)+
  theme_minimal()

```




