---
title: "State Level Analysis of Risk for Asthma and Covid-19"
author: "Esteban Correa"
date: "April 21, 2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
library(tidyverse)
library(SpatialEpi)
library(raster)
library(rgeos)
library(rgdal)
library(sf)
library(leaflet)
library(RColorBrewer)
library(ggthemes)
library(ggplot2)
library(lubridate)
library(classInt)
library(brinla)
library(INLA)
library(spdep)
library(brms)
library(brinla)
library(performance)

covid19<-read.csv("data/covid04-24-2020.csv") %>%
  dplyr::select(FIPS,Covid.Confirmed,Covid.Deaths) %>%
  mutate(covidCFR=Covid.Deaths*100/Covid.Confirmed)
clrdDeaths<-read.csv("data/chroniclowerRespDiseaseDeaths.csv") %>%
  mutate(clrdSMR=clrd.Deaths*100000/clrd.Population)
  # dplyr::select(County.Code,clrd.Deaths,clrd.Population,Crude.Rate) 
cdi<-read.csv("data/CDIByState.csv")
pm25<-read.csv("data/pm2.5byCounty.csv")
demog<-read.csv("data/ACS5Years2018.csv")

uscounty<-readOGR("data/SVI","SVI2018_US_county")
uscounty@data<-uscounty@data %>%
  dplyr::select(FIPS,ST,ST_ABBR,E_TOTPOP,EP_POV)

# ICD-10 Codes: J40-J47 (Chronic lower respiratory diseases)
```

# Covid19 at April 20-2020, Asthma deaths from 2010-2017

```{r,include=FALSE}
clrdDeaths$clrd.Deaths<-as.numeric(as.character(clrdDeaths$clrd.Deaths))
clrdDeaths$clrd.Population<-as.numeric(as.character(clrdDeaths$clrd.Population))
clrdDeaths$Crude.Rate<-as.numeric(as.character(clrdDeaths$Crude.Rate))
clrdDeaths$fips<-sprintf("%05d", clrdDeaths$County.Code)
clrdDeaths$clrd.Deaths[is.na(clrdDeaths$clrd.Deaths)]<-0

covid19$id<-sprintf("%05d", covid19$FIPS) 
covid19$Covid.Deaths[is.na(covid19$Covid.Deaths)]<-0


# age groups
# DP05_0008PE 1519
# DP05_0009PE 2024
# DP05_0010PE 2534
# DP05_0011PE 3544
# DP05_0012PE 4554
# DP05_0013PE 5559
# DP05_0014PE 6064




demog$fips<-substring(demog$GEO_ID,10,14)
demog<-demog%>% 
  dplyr::select(fips,DP05_0037PE,DP05_0038PE,DP05_0071PE,DP05_0002PE,DP05_0003PE,DP05_0024PE,DP05_0008PE,
DP05_0009PE,
DP05_0010PE,
DP05_0011PE,
DP05_0012PE,
DP05_0013PE,
DP05_0014PE,) %>% 
  mutate(a1524=DP05_0008PE+DP05_0009PE,
         a5564=DP05_0013PE+DP05_0014PE)

colnames(demog)<-c("fips","whitePE","blackPE","latinoPE","malePE","femalePE",
                   "over65PE","a1519","a2024","a2534","a3544","a4554","a5559","a6064","a1524","a5564")


pm25$fips<-sprintf("%05d", pm25$fips)
pm25$pm25avg<-rowMeans(pm25[,2:20],na.rm = T)
pm25<-pm25%>% 
  dplyr::select(fips,pm25avg)



```



<!-- # County level PM 2.5 -->

```{r}
uscounty <- merge(uscounty, cdi,by.x="ST_ABBR",by.y="ST_ABBR") #Chronic diseases
uscounty <- merge(uscounty, covid19,by.x="FIPS",by.y="id") #covidDeaths
uscounty <- merge(uscounty, clrdDeaths,by.x="FIPS",by.y="fips") #asthmaDe
uscounty <- merge(uscounty, demog,by.x="FIPS",by.y="fips") #demog
uscounty <- merge(uscounty, pm25,by.x="FIPS",by.y="fips") #pm25

uscounty$clrd.Deaths[is.na(uscounty$clrd.Deaths)]<-0
uscounty$Covid.Deaths[is.na(uscounty$Covid.Deaths)]<-0
#removing alaska y puerto rico
uscounty<-uscounty[uscounty$ST!="02",]
uscounty<-uscounty[uscounty$ST!="15",]
uscounty<-uscounty[uscounty$ST!="36",]
uscounty<-uscounty[uscounty$FIPS!="35039",]
head(uscounty,100)
```




```{r}

# head(uscounty,100)
uscounty$E_TOTPOP<-as.numeric(as.character(uscounty$E_TOTPOP))
uscounty$Easthma <- expected(uscounty$E_TOTPOP, uscounty$clrd.Deaths, 1)
uscounty$Ecovid <- expected(uscounty$E_TOTPOP, uscounty$Covid.Deaths, 1)

uscounty$SIR.asthma <- uscounty$clrd.Deaths/uscounty$Easthma
uscounty$SIR.covid <- uscounty$Covid.Deaths/uscounty$Ecovid
head(uscounty)
```

# ```{r}
# hist(log(uscounty$covidCFR),64)
# 
# hist(uscounty$Covid.Deaths,64)
# hist(uscounty$clrd.Deaths,64)
# hist(uscounty$whitePE,64)
# hist(uscounty$over65PE,64)
# plot(uscounty$Covid.Deaths, uscounty$clrd.Deaths)
# 
# plot(uscounty$Covid.Deaths, uscounty$smokingPrev)
# plot(uscounty$Covid.Deaths, uscounty$avgObesityPrev)
# 
# plot(uscounty$Covid.Deaths, uscounty$over65PE)
# plot(uscounty$Covid.Deaths, uscounty$blackPE)
# plot(uscounty$Covid.Deaths, uscounty$whitePE)
# ```

# Covariates

EP_MINRTY = Percentage minority (all persons except white, nonHispanic) estimate E_PCI Per capita income estimate, 2014-2018 ACS.
EP_CROWD = At household level (occupied housing units), more people than rooms estimate, 2014-2018 ACS
EP_AGE65  = percentage of people age 65 over
EP_POV = Percentage of people living under poverty
EP_CROWD = Percentage of occupied housing units with more people than rooms estimate

<!-- # Asthma deaths 1 -->

<!-- $Asthma_{Deaths}\sim covidMortality_{100k} + pm25_{2018} + EP\_POV_{\%} + EP\_AGE65_{\%} + EP\_MINRTY_{\%} + EP\_CROWD_{\%} + (1|STATE)$ -->

<!-- ```{r,include = FALSE} -->
<!-- sirMap<-uscounty -->
<!-- sirMap$E_PCI<-as.numeric(sirMap$E_PCI) -->
<!-- sirMap$E_MINRTY<-as.numeric(sirMap$E_MINRTY) -->
<!-- sirMap$EP_AGE65<-as.numeric(sirMap$EP_AGE65) -->
<!-- sirMap$Covid.Confirmed<-as.numeric(sirMap$Covid.Confirmed) -->
<!-- # hist(sirMap$Covid.Confirmed,128) -->
<!-- sirMap$Covid.Confirmed.p100k<-sirMap$Covid.Confirmed*100/sirMap$E_TOTPOP -->
<!-- sirMap$Covid.Deaths.p100k<-sirMap$Covid.Deaths*100/sirMap$E_TOTPOP -->
<!-- sirMap$clrd.Deaths.p100k<-sirMap$clrd.Deaths*100/sirMap$E_TOTPOP -->

<!-- sirMap<-sirMap[sirMap$EP_POV>=0,] -->
<!-- sirMap<-sirMap[sirMap$EP_AGE65>=0,] -->
<!-- sirMap<-sirMap[sirMap$EP_MINRTY>=0,] -->
<!-- sirMap<-sirMap[sirMap$EP_CROWD>=0,] -->
<!-- # nb <- poly2nb(sirMap) -->
<!-- # nb2INLA("us.adj", nb) -->
<!-- # g <- inla.read.graph(filename = "us.adj") -->
<!-- # sirMap$re_u <- 1:nrow(sirMap@data) -->
<!-- sirMap$re_v <- 1:nrow(sirMap@data) -->


<!-- f1<-clrd.Deaths ~  Covid.Deaths.p100k + pm25 + EP_POV + EP_AGE65 + EP_MINRTY + EP_CROWD + f(STATE,model = "iid") + f(re_v, model = "iid") -->
<!-- inla.m1 <- inla( f1, -->
<!--                  family = "poisson",data = sirMap@data,E = Easthma, -->
<!--                  num.threads = 8,verbose = F, -->
<!--                  control.predictor = list(compute = TRUE), -->
<!--                  control.compute = list(dic = TRUE,waic=T,cpo=TRUE)) -->

<!-- sirMap$RRasth <- inla.m1$summary.fitted.values[, "mean"] -->
<!-- sirMap$LLasth <- inla.m1$summary.fitted.values[, "0.025quant"] -->
<!-- sirMap$ULasth <- inla.m1$summary.fitted.values[, "0.975quant"] -->
<!-- ``` -->


<!-- ```{r,echo=FALSE} -->

<!-- summary(inla.m1) -->
<!-- round(exp(inla.m1$summary.fixed[,c(1,3,5)]),2) -->

<!-- # bri.Pois.resid(inla.m1) -->
<!-- #check random effects -->
<!-- # bri.hyperpar.summary(inla.m1) -->
<!-- # check poisson assumption where u(x)=var(x) -->
<!-- # bri.hyperpar.plot(inla.m1) -->


<!-- tab <- inla.m1$summary.random$STATE -->
<!-- names(tab)[c(4, 6)] <- c("IClower", "ICupper") -->
<!-- tab<-tab[order(tab$ID),] -->
<!-- tab$idx<-nrow(tab):1 -->
<!-- tab$mean_exp<-exp(tab$mean) -->
<!-- tab<-tab[!is.infinite(tab$mean_exp),] -->
<!-- tab$IClower_exp<-exp(tab$IClower) -->
<!-- tab$ICupper_exp<-exp(tab$ICupper) -->
<!-- ID<-tab$ID -->
<!-- ggplot(data = tab, aes(x = mean_exp, y = idx)) + -->
<!--   geom_point() + xlab(expression(u[p])) + -->
<!--   scale_x_continuous(breaks = NULL)+ -->
<!--   scale_y_continuous("STATE", breaks=nrow(tab):1,labels=c(ID)) + -->
<!--   geom_errorbarh(aes(xmin = IClower_exp, xmax = ICupper_exp) ) + -->
<!--   geom_vline (xintercept = 1, linetype = 2)+ -->
<!--   theme_fivethirtyeight() -->
<!-- ``` -->


<!-- ```{r,echo=FALSE} -->

<!-- brks<-c(min(sirMap$RRasth),1.0,max(sirMap$RRasth)) -->
<!-- ggplot() + -->
<!--   geom_sf(data=st_as_sf(sirMap),aes(fill = RRasth)) + -->
<!--   scale_fill_gradient2( -->
<!--     midpoint = 1, low = "blue", mid = "white", high = "red", -->
<!--      breaks = brks, -->
<!--     labels = round(brks,1) -->
<!--   ) + -->
<!--   ggtitle("Relative Risk Asthma risk in Covid-19 Pandemic")+ -->
<!--   coord_sf(crs = mycrs)+ -->
<!--   theme_minimal() -->

<!-- ``` -->

<!-- # Asthma deaths 2 -->

Do the case fatality rate of covid19 increase based on long term exposure and asthma prevalence?

```{r}
sirMap<-uscounty
# sirMap$E_PCI<-as.numeric(sirMap$E_PCI)
# sirMap$E_MINRTY<-as.numeric(sirMap$E_MINRTY)
# sirMap$EP_AGE65<-as.numeric(sirMap$EP_AGE65)
sirMap$Covid.Confirmed<-as.numeric(sirMap$Covid.Confirmed)
# hist(sirMap$Covid.Confirmed,128)
sirMap$Covid.Confirmed.p100k<-sirMap$Covid.Confirmed*100/sirMap$E_TOTPOP
sirMap$Covid.Deaths[is.na(sirMap$Covid.Deaths)]<-0
sirMap$Covid.Deaths.p100k<-sirMap$Covid.Deaths*100/sirMap$E_TOTPOP
sirMap$clrd.Deaths[is.na(sirMap$clrd.Deaths)]<-0
sirMap$clrd.Deaths.p100k<-sirMap$clrd.Deaths*100/sirMap$E_TOTPOP


# nb <- poly2nb(sirMap)
# nb2INLA("us.adj", nb)
# g <- inla.read.graph(filename = "us.adj")
# sirMap$re_u <- 1:nrow(sirMap@data)
sirMap$re_v <- 1:nrow(sirMap@data)

f1<- covidCFR ~  asthmaPrev + a1524+a2534+a3544+a4554+a5564+over65PE + pm25avg + EP_POV + blackPE+latinoPE + malePE+femalePE + avgObesityPrev + smokingCOPD +  f(ST_ABBR,model = "iid")
inla.m2 <- inla( f1,
                 # family = "zeroinflatedpoisson1",
                 # family = "poisson",
                 family = "gaussian",
                 data = sirMap@data,
                 num.threads = 16,verbose = F,
                 control.predictor = list(compute = TRUE),
                 control.compute = list(dic = TRUE,waic=T,cpo=TRUE))

sirMap$RRasth2 <- inla.m2$summary.fitted.values[, "mean"]
sirMap$LLasth2 <- inla.m2$summary.fitted.values[, "0.025quant"]
sirMap$ULasth2 <- inla.m2$summary.fitted.values[, "0.975quant"]
summary(inla.m2)
round(exp(inla.m2$summary.fixed[,c(1,3,5)]),2)
bri.hyperpar.plot(inla.m2)
# bri.Pois.resid(inla.m2)
tab <- inla.m2$summary.random$ST_ABBR
names(tab)[c(4, 6)] <- c("IClower", "ICupper")
tab<-tab[order(tab$ID),]
tab$idx<-nrow(tab):1
tab$mean_exp<-exp(tab$mean)
tab<-tab[!is.infinite(tab$mean_exp),]
tab$IClower_exp<-exp(tab$IClower)
tab$ICupper_exp<-exp(tab$ICupper)
ID<-tab$ID
ggplot(data = tab, aes(x = mean_exp, y = idx)) +
  geom_point() + xlab(expression(u[p])) +
  scale_x_continuous(breaks = NULL)+
  scale_y_continuous("STATE", breaks=nrow(tab):1,labels=c(ID)) +
  geom_errorbarh(aes(xmin = IClower_exp, xmax = ICupper_exp) ) +
  geom_vline (xintercept = 1, linetype = 2)+
  theme_fivethirtyeight()

# 0.105^2/(0.105^2+0.2625^2)
# [1] 0.137931
```

---------------------------------------------------------------------------

Do the mortality of chronic lower respiratory diseases increase with covid19?

```{r}
sirMap<-uscounty
# sirMap$E_PCI<-as.numeric(sirMap$E_PCI)
# sirMap$E_MINRTY<-as.numeric(sirMap$E_MINRTY)
# sirMap$EP_AGE65<-as.numeric(sirMap$EP_AGE65)
sirMap$Covid.Confirmed<-as.numeric(sirMap$Covid.Confirmed)
# hist(sirMap$Covid.Confirmed,128)
sirMap$Covid.Confirmed.p100k<-sirMap$Covid.Confirmed*100/sirMap$E_TOTPOP
sirMap$Covid.Deaths[is.na(sirMap$Covid.Deaths)]<-0
sirMap$Covid.Deaths.p100k<-sirMap$Covid.Deaths*100/sirMap$E_TOTPOP
sirMap$clrd.Deaths[is.na(sirMap$clrd.Deaths)]<-0
sirMap$clrd.Deaths.p100k<-sirMap$clrd.Deaths*100/sirMap$E_TOTPOP


# nb <- poly2nb(sirMap)
# nb2INLA("us.adj", nb)
# g <- inla.read.graph(filename = "us.adj")
# sirMap$re_u <- 1:nrow(sirMap@data)
sirMap$re_v <- 1:nrow(sirMap@data)

f1<-clrd.Deaths ~  Covid.Deaths + a1524+a2534+a3544+a4554+a5564+over65PE + pm25avg + EP_POV + blackPE+latinoPE + malePE+femalePE + avgObesityPrev + smokingCOPD + asthmaPrev + f(ST_ABBR,model = "iid") + f(re_v, model = "iid")
inla.m2 <- inla( f1,
                 # family = "zeroinflatedpoisson1",
                 family = "poisson",
                 data = sirMap@data,E = Easthma,
                 num.threads = 8,verbose = F,
                 control.predictor = list(compute = TRUE),
                 control.compute = list(dic = TRUE,waic=T,cpo=TRUE))

sirMap$RRasth2 <- inla.m2$summary.fitted.values[, "mean"]
sirMap$LLasth2 <- inla.m2$summary.fitted.values[, "0.025quant"]
sirMap$ULasth2 <- inla.m2$summary.fitted.values[, "0.975quant"]
summary(inla.m2)
round(exp(inla.m2$summary.fixed[,c(1,3,5)]),2)
bri.hyperpar.plot(inla.m2)
bri.Pois.resid(inla.m2)
tab <- inla.m2$summary.random$ST_ABBR
names(tab)[c(4, 6)] <- c("IClower", "ICupper")
tab<-tab[order(tab$ID),]
tab$idx<-nrow(tab):1
tab$mean_exp<-exp(tab$mean)
tab<-tab[!is.infinite(tab$mean_exp),]
tab$IClower_exp<-exp(tab$IClower)
tab$ICupper_exp<-exp(tab$ICupper)
ID<-tab$ID
ggplot(data = tab, aes(x = mean_exp, y = idx)) +
  geom_point() + xlab(expression(u[p])) +
  scale_x_continuous(breaks = NULL)+
  scale_y_continuous("STATE", breaks=nrow(tab):1,labels=c(ID)) +
  geom_errorbarh(aes(xmin = IClower_exp, xmax = ICupper_exp) ) +
  geom_vline (xintercept = 1, linetype = 2)+
  theme_fivethirtyeight()

# 0.105^2/(0.105^2+0.2625^2)
# [1] 0.137931
```

               mean 0.025quant 0.975quant
(Intercept)    1.01       0.78       1.30
Covid.Deaths   1.00       1.00       1.00
a1524          0.97       0.97       0.97
a2534          0.99       0.98       1.00
a3544          0.98       0.97       0.99
a4554          0.99       0.98       1.00
a5564          0.99       0.98       1.00
over65PE       1.03       1.02       1.03
pm25avg        1.01       1.00       1.01
EP_POV         1.02       1.02       1.02
blackPE        0.99       0.99       0.99
latinoPE       0.99       0.99       0.99
malePE         1.00       1.00       1.01
femalePE       1.00       0.99       1.00
avgObesityPrev 1.01       1.00       1.02
smokingCOPD    1.00       1.00       1.01
asthmaPrev     1.00       0.98       1.02

Podemos concluir que no hay suficiene evidencia para que aumente el riesgo debido a covid19 o polucion


# brms



```{r}
brmsDF<-as.data.frame(uscounty)

brm.c19 <- brm(Covid.Deaths ~ offset(log(Ecovid)) + clrdSMR + a1524+a2534+a3544+a4554+a5564+over65PE + pm25avg + EP_POV + blackPE+latinoPE + malePE+ femalePE + avgObesityPrev  + smokingPrev + asthmaPrev + htaPrev + chdPrev + (1|ST_ABBR), 
              data = brmsDF, 
              family = "poisson",chains=6,
              # family = poisson(),
              iter = 10000, cores  = 16,seed = 2020)  
summary(brm.c19)
round(exp(summary(brm.c19)$fixed[,c(1,3,4)]),2)

pp_check(brm.c19,nsamples = nrow(brmsDF)/1)
plot(brm.c19, N = 2, ask = FALSE)
plot(brm.c19, points = TRUE,ask=FALSE)
# names(brm.bmi)
# ## extract fitted values
fitted_values <- fitted(brm.c19)
```

Group-Level Effects: 
~ST_ABBR (Number of levels: 48) 
              Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
sd(Intercept)     1.02      0.11     0.87     1.28 1.59       10       33

Population-Level Effects: 
               Estimate Est.Error  l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
Intercept      -6675.54  13392.52 -36861.39  7620.67 2.55        7       24
clrdSMR           -0.01      0.00     -0.01    -0.01 1.09       42      138
a1524              0.02      0.01      0.01     0.03 1.47       12       18
a2534              0.06      0.01      0.04     0.07 1.61       10       21
a3544              0.08      0.01      0.05     0.10 1.51       11       27
a4554              0.32      0.02      0.29     0.36 1.75        9       29
a5564             -0.11      0.01     -0.13    -0.09 1.49       11       38
over65PE           0.16      0.01      0.15     0.17 1.47       12       25
pm25avg            0.09      0.01      0.08     0.10 1.33       15       51
EP_POV             0.02      0.00      0.02     0.03 1.23       19       44
blackPE            0.02      0.00      0.02     0.02 1.21       20       81
latinoPE           0.02      0.00      0.02     0.02 1.18       23       53
malePE            66.62    133.93    -76.33   368.49 2.55        7       24
femalePE          66.76    133.93    -76.19   368.63 2.55        7       24
avgObesityPrev     0.07      0.06     -0.01     0.20 1.69       10       44
smokingPrev       -0.05      0.10     -0.26     0.08 2.09        8       23
asthmaPrev         0.02      0.12     -0.19     0.26 1.93        8       22
htaPrev           -0.20      0.04     -0.29    -0.13 1.41       13       47
chdPrev            0.71      0.37     -0.06     1.32 2.45        7       19

Samples were drawn using sampling(NUTS). For each parameter, Bulk_ESS
and Tail_ESS are effective sample size measures, and Rhat is the potential
scale reduction factor on split chains (at convergence, Rhat = 1).
Parts of the model have not converged (some Rhats are > 1.05). Be careful when analysing the results! We recommend running more iterations and/or setting stronger priors.                   Estimate l-95% CI      u-95% CI
Intercept      0.000000e+00     0.00           Inf
clrdSMR        9.900000e-01     0.99  9.900000e-01
a1524          1.020000e+00     1.01  1.030000e+00
a2534          1.060000e+00     1.04  1.070000e+00
a3544          1.080000e+00     1.06  1.100000e+00
a4554          1.380000e+00     1.34  1.430000e+00
a5564          9.000000e-01     0.88  9.200000e-01
over65PE       1.170000e+00     1.16  1.190000e+00
pm25avg        1.090000e+00     1.08  1.110000e+00
EP_POV         1.020000e+00     1.02  1.030000e+00
blackPE        1.020000e+00     1.02  1.020000e+00
latinoPE       1.020000e+00     1.02  1.020000e+00
malePE         8.545021e+28     0.00 1.082402e+160
femalePE       9.856587e+28     0.00 1.242767e+160
avgObesityPrev 1.070000e+00     0.99  1.220000e+00
smokingPrev    9.500000e-01     0.77  1.090000e+00
asthmaPrev     1.020000e+00     0.83  1.300000e+00
htaPrev        8.200000e-01     0.75  8.800000e-01
chdPrev        2.030000e+00     0.95  3.750000e+00

# INLA Version

```{r}
sirMap<-uscounty
# sirMap$E_PCI<-as.numeric(sirMap$E_PCI)
# sirMap$E_MINRTY<-as.numeric(sirMap$E_MINRTY)
# sirMap$EP_AGE65<-as.numeric(sirMap$EP_AGE65)
sirMap$Covid.Confirmed<-as.numeric(sirMap$Covid.Confirmed)
# hist(sirMap$Covid.Confirmed,128)
sirMap$Covid.Confirmed.p100k<-sirMap$Covid.Confirmed*100/sirMap$E_TOTPOP
sirMap$Covid.Deaths[is.na(sirMap$Covid.Deaths)]<-0
sirMap$Covid.Deaths.p100k<-sirMap$Covid.Deaths*100/sirMap$E_TOTPOP
sirMap$clrd.Deaths[is.na(sirMap$clrd.Deaths)]<-0
sirMap$clrd.Deaths.p100k<-sirMap$clrd.Deaths*100/sirMap$E_TOTPOP


# nb <- poly2nb(sirMap)
# nb2INLA("us.adj", nb)
# g <- inla.read.graph(filename = "us.adj")
# sirMap$re_u <- 1:nrow(sirMap@data)
sirMap$re_v <- 1:nrow(sirMap@data)

f1<-Covid.Deaths ~  clrdSMR + a1524+a2534+a3544+a4554+a5564+over65PE + pm25avg + EP_POV + blackPE+latinoPE + malePE+ femalePE + avgObesityPrev  + smokingPrev + asthmaPrev + htaPrev + chdPrev + f(ST_ABBR,model = "iid") + f(re_v, model = "iid")
inla.cv19 <- inla( f1,
                 # family = "zeroinflatedpoisson1",
                 family = "poisson",
                 data = sirMap@data,E = Easthma,
                 num.threads = 8,verbose = F,
                 control.predictor = list(compute = TRUE),
                 control.compute = list(dic = TRUE,waic=T,cpo=TRUE))

sirMap$RRcv19<- inla.cv19$summary.fitted.values[, "mean"]
sirMap$LLcv19 <- inla.cv19$summary.fitted.values[, "0.025quant"]
sirMap$ULcv19 <- inla.cv19$summary.fitted.values[, "0.975quant"]
summary(inla.cv19)
round(exp(inla.cv19$summary.fixed[,c(1,3,5)]),2)
bri.hyperpar.plot(inla.cv19)
bri.Pois.resid(inla.cv19)
tab <- inla.m2$summary.random$ST_ABBR
names(tab)[c(4, 6)] <- c("IClower", "ICupper")
tab<-tab[order(tab$ID),]
tab$idx<-nrow(tab):1
tab$mean_exp<-exp(tab$mean)
tab<-tab[!is.infinite(tab$mean_exp),]
tab$IClower_exp<-exp(tab$IClower)
tab$ICupper_exp<-exp(tab$ICupper)
ID<-tab$ID
ggplot(data = tab, aes(x = mean_exp, y = idx)) +
  geom_point() + xlab(expression(u[p])) +
  scale_x_continuous(breaks = NULL)+
  scale_y_continuous("STATE", breaks=nrow(tab):1,labels=c(ID)) +
  geom_errorbarh(aes(xmin = IClower_exp, xmax = ICupper_exp) ) +
  geom_vline (xintercept = 1, linetype = 2)+
  theme_fivethirtyeight()
```


# BRMS sin random slopes

```{r}
brmsDF<-as.data.frame(uscounty)

brm.fl3 <- brm(Covid.Deaths ~ offset(log(Ecovid)) + clrdSMR + a1524+a2534+a3544+a4554+a5564+over65PE + pm25avg + EP_POV + blackPE+latinoPE + malePE+ femalePE + avgObesityPrev  + smokingPrev + asthmaPrev + (1|ST_ABBR), 
              data = brmsDF, 
              family = "zero_inflated_poisson",
              # family = poisson(),
              iter = 3000, cores  = 8,seed = 2020)  
summary(brm.fl3)
round(exp(summary(brm.fl3)$fixed[,c(1,3,4)]),2)

pp_check(brm.fl3,nsamples = nrow(brmsDF)/1)
plot(brm.fl3, N = 2, ask = FALSE)
plot(brm.fl3, points = TRUE,ask=FALSE)
# names(brm.bmi)
# ## extract fitted values
fitted_values <- fitted(brm.fl3)
```

clrdSMR         9.900000e-01     0.99 9.900000e-01
a1524           1.010000e+00     1.00 1.030000e+00
a2534           1.050000e+00     1.04 1.070000e+00
a3544           1.140000e+00     1.10 1.170000e+00
a4554           1.350000e+00     1.31 1.390000e+00
a5564           9.300000e-01     0.91 9.500000e-01
over65PE        1.170000e+00     1.16 1.190000e+00
pm25avg         1.070000e+00     1.06 1.090000e+00
EP_POV          1.030000e+00     1.03 1.040000e+00
blackPE         1.020000e+00     1.02 1.020000e+00
latinoPE        1.020000e+00     1.01 1.020000e+00
malePE          1.000000e-02     0.00 2.839830e+22
femalePE        1.000000e-02     0.00 3.228679e+22
avgObesityPrev  9.900000e-01     0.86 1.130000e+00
smokingPrev     1.020000e+00     0.86 1.120000e+00
asthmaPrev      9.100000e-01     0.76 1.140000e+00


# BRMS sin random slopes scale

```{r}
brmsDF<-as.data.frame(uscounty)

brm.fl3 <- brm(Covid.Deaths ~ offset(log(Ecovid)) + clrdSMR + scale(a1524)+scale(a2534)+scale(a3544)+scale(a4554)+scale(a5564)+scale(over65PE) + pm25avg + scale(EP_POV) + scale(blackPE)+scale(latinoPE) + scale(malePE)+ scale(femalePE) + scale(avgObesityPrev)  + scale(smokingPrev) + scale(asthmaPrev) + (1|ST_ABBR), 
              data = brmsDF, 
              family = "zero_inflated_poisson",chains = 6,
              # family = poisson()
              iter = 4000, cores  = 8,seed = 2020)  
summary(brm.fl3)
round(exp(summary(brm.fl3)$fixed[,c(1,3,4)]),2)

pp_check(brm.fl3,nsamples = nrow(brmsDF)/1)
plot(brm.fl3, N = 2, ask = FALSE)
plot(brm.fl3, points = TRUE,ask=FALSE)
# names(brm.bmi)
# ## extract fitted values
fitted_values <- fitted(brm.fl3)
```

Intercept               0.51     0.40  6.300000e-01
clrdSMR                 0.99     0.99  9.900000e-01
scalea1524              1.05     1.00  1.100000e+00
scalea2534              1.12     1.10  1.150000e+00
scalea3544              1.21     1.15  1.270000e+00
scalea4554              1.57     1.52  1.630000e+00
scalea5564              0.85     0.80  9.000000e-01
scaleover65PE           2.09     1.95  2.190000e+00
pm25avg                 1.07     1.06  1.090000e+00
scaleEP_POV             1.24     1.19  1.290000e+00
scaleblackPE            1.28     1.24  1.340000e+00
scalelatinoPE           1.25     1.21  1.300000e+00
scalemalePE             0.00     0.00 5.786269e+104
scalefemalePE           0.00     0.00 7.887653e+104
scaleavgObesityPrev     0.95     0.78  1.250000e+00
scalesmokingPrev        0.91     0.69  1.100000e+00
scaleasthmaPrev         0.88     0.70  1.050000e+00



# Covid deaths

$Covid_{Deaths}\sim asthmaMortality_{100k} + pm25_{2018} + EP\_POV_{\%} + EP\_AGE65_{\%} + EP\_MINRTY_{\%} + EP\_CROWD_{\%} + (1|STATE)$

<!-- ```{r,include=FALSE} -->

<!-- # plot(sirMap$clrd.Deaths.rate) -->

<!-- f1<- Covid.Deaths ~  sirMap$clrd.Deaths.p100k + pm25avg + EP_POV + EP_AGE65 + EP_MINRTY + EP_CROWD + f(STATE,model = "iid") + f(re_v, model = "iid") -->
<!-- inla.m1 <- inla( f1, -->
<!--                  family = "zeroinflatedpoisson1", -->
<!--                  data = sirMap@data,E = Easthma, -->
<!--                  num.threads = 8,verbose = F, -->
<!--                  control.predictor = list(compute = TRUE), -->
<!--                  control.compute = list(dic = TRUE,waic=T,cpo=TRUE)) -->

<!-- sirMap$RRcov <- inla.m1$summary.fitted.values[, "mean"] -->
<!-- sirMap$LLcov <- inla.m1$summary.fitted.values[, "0.025quant"] -->
<!-- sirMap$ULcov <- inla.m1$summary.fitted.values[, "0.975quant"] -->
<!-- ``` -->


Do the covid19 death risk increase with covid19?

```{r}
sirMap<-uscounty

sirMap$Covid.Confirmed<-as.numeric(sirMap$Covid.Confirmed)
sirMap$Covid.Confirmed.p100k<-sirMap$Covid.Confirmed*100/sirMap$E_TOTPOP
sirMap$Covid.Deaths[is.na(sirMap$Covid.Deaths)]<-0
sirMap$Covid.Deaths.p100k<-sirMap$Covid.Deaths*100/sirMap$E_TOTPOP
sirMap$clrd.Deaths[is.na(sirMap$clrd.Deaths)]<-0
sirMap$clrd.Deaths.p100k<-sirMap$clrd.Deaths*100/sirMap$E_TOTPOP

sirMap$re_v <- 1:nrow(sirMap@data)

f1<-Covid.Deaths ~  clrdSMR + a1524+a2534+a3544+a4554+a5564+over65PE + pm25avg + EP_POV + blackPE+latinoPE + malePE+femalePE + avgObesityPrev + smokingCOPD + diabetesPrev + ckdPrev + asthmaPrev+ f(ST_ABBR,model = "iid") + f(re_v, model = "iid")
inla.m2 <- inla( f1,
                 family = "zeroinflatedpoisson1",
                 # family = "poisson",
                 data = sirMap@data,E = Ecovid,
                 num.threads = 8,verbose = F,
                 control.predictor = list(compute = TRUE),
                 control.compute = list(dic = TRUE,waic=T,cpo=TRUE))

sirMap$RRasth2 <- inla.m2$summary.fitted.values[, "mean"]
sirMap$LLasth2 <- inla.m2$summary.fitted.values[, "0.025quant"]
sirMap$ULasth2 <- inla.m2$summary.fitted.values[, "0.975quant"]

summary(inla.m2)
round(exp(inla.m2$summary.fixed[,c(1,3,5)]),2)
bri.hyperpar.plot(inla.m2)
bri.Pois.resid(inla.m2)

tab <- inla.m2$summary.random$ST_ABBR
names(tab)[c(4, 6)] <- c("IClower", "ICupper")
tab<-tab[order(tab$ID),]
tab$idx<-nrow(tab):1
tab$mean_exp<-exp(tab$mean)
tab<-tab[!is.infinite(tab$mean_exp),]
tab$IClower_exp<-exp(tab$IClower)
tab$ICupper_exp<-exp(tab$ICupper)
ID<-tab$ID
ggplot(data = tab, aes(x = mean_exp, y = idx)) +
  geom_point() + xlab(expression(u[p])) +
  scale_x_continuous(breaks = NULL)+
  scale_y_continuous("STATE", breaks=nrow(tab):1,labels=c(ID)) +
  geom_errorbarh(aes(xmin = IClower_exp, xmax = ICupper_exp) ) +
  geom_vline (xintercept = 1, linetype = 2)+
  theme_fivethirtyeight()

# 0.9^2/(0.9^2+1.1^2)
# [1] 0.4009901  40% is explained by the state level 
```

                             mean 0.025quant 0.975quant
(Intercept)    0.89       0.68       1.16
clrdSMR        1.00       0.99       1.00
a1524          0.98       0.97       0.99
a2534          0.99       0.98       1.01
a3544          1.00       0.98       1.02
a4554          1.00       0.99       1.02
a5564          0.99       0.98       1.01
over65PE       1.00       0.99       1.01
pm25avg        1.03       1.00       1.05
EP_POV         0.99       0.98       1.00
blackPE        1.02       1.02       1.03
latinoPE       1.00       1.00       1.01
malePE         1.00       0.99       1.00
femalePE       1.00       0.99       1.00
avgObesityPrev 1.00       0.99       1.00
smokingCOPD    1.00       0.99       1.00
diabetesPrev   1.00       0.99       1.00
ckdPrev        0.97       0.89       1.05
asthmaPrev     0.99       0.96       1.02

Podemos concluir que una larga exposicion aumenta en un 3% el riesgo de muerte por cada 


Analizado con zeroinflated negative binomial

```{r}
library("lme4")
library(NBZIMM)
# main analysis
glmm.zinb.off = glmm.zinb(Covid.Deaths ~  clrd.Deaths + a1524+a2534+a3544+a4554+a5564+over65PE + pm25avg + EP_POV + blackPE+latinoPE + malePE+femalePE + avgObesityPrev + smokingCOPD + diabetesPrev + ckdPrev + asthmaPrev + offset(log(Ecovid)), data = (sirMap@data))

fixed(glmm.zinb.off)$dist[2,1]
fixed(glmm.zinb.off)$dist[2,1] - 1.96*fixed(glmm.zinb.off)$dist[2,2]
fixed(glmm.zinb.off)$dist[2,1] + 1.96*fixed(glmm.zinb.off)$dist[2,2]
fixed(glmm.zinb.off)$dist[2,3]
```



```{r}
sirMap<-uscounty

sirMap$Covid.Confirmed<-as.numeric(sirMap$Covid.Confirmed)
sirMap$Covid.Confirmed.p100k<-sirMap$Covid.Confirmed*100/sirMap$E_TOTPOP
sirMap$Covid.Deaths[is.na(sirMap$Covid.Deaths)]<-0
sirMap$Covid.Deaths.p100k<-sirMap$Covid.Deaths*100/sirMap$E_TOTPOP
sirMap$clrd.Deaths[is.na(sirMap$clrd.Deaths)]<-0
sirMap$clrd.Deaths.p100k<-sirMap$clrd.Deaths*100/sirMap$E_TOTPOP

sirMap$re_v <- 1:nrow(sirMap@data)

f2<-Covid.Deaths ~  clrd.Deaths + pm25avg + EP_POV + blackPE+latinoPE + femalePE + over65PE + avgObesityPrev + smokingCOPD + diabetesPrev + ckdPrev + f(re_v, model = "iid")
inla.m3 <- inla( f2,
                 # family = "zeroinflatedpoisson1",
                 family = "poisson",
                 data = sirMap@data,E = Ecovid,
                 num.threads = 8,verbose = F,
                 control.predictor = list(compute = TRUE),
                 control.compute = list(dic = TRUE,waic=T,cpo=TRUE))

sirMap$RRcov2 <- inla.m3$summary.fitted.values[, "mean"]
sirMap$LLcov2 <- inla.m3$summary.fitted.values[, "0.025quant"]
sirMap$ULcov2 <- inla.m3$summary.fitted.values[, "0.975quant"]

summary(inla.m3)
round(exp(inla.m3$summary.fixed[,c(1,3,5)]),2)
bri.hyperpar.plot(inla.m3)
bri.Pois.resid(inla.m3)

```


```{r,echo=FALSE}

brks<-c(min(sirMap$RRasth2),1.0,max(sirMap$RRasth2))
ggplot() +
  geom_sf(data=st_as_sf(sirMap),aes(fill = RRasth2)) +
  scale_fill_gradient2(
    midpoint = 1, low = "blue", mid = "white", high = "red",
     breaks = brks,
    labels = round(brks,1)
  ) +
  ggtitle("Relative Risk for Covid-19 Pandemic")+
  # coord_sf(crs = mycrs)+
  theme_minimal()

```




